---
title: ""
author: "Sandra Cortes, Alan Domínguez, María Jose Ojeda"
date: "7/23/2021"
output: html_document
---

## CAYAFAM 1

The objective of this study was to analyze the association between air pollution and cardiorespiratory variables, considering the effect of physical condition, in OA that live in areas with high levels of air pollution in the Metropolitan Region of Chile.

```{r Libraries, include=FALSE}
library(tidyverse)
library(skimr)
library(haven)
library(readxl)
library(patchwork)
library(gmodels)
library(summarytools)
library(ggpubr)
library(ggthemes)
library(DataExplorer)

```

To charge the database from SPSS we use the function read_sav from the library(haven) for the initial database, and the function read_excel from the library(readxl).

```{r Loading Database, echo = FALSE}
# Base de datos utilizada (contiene la mayoria de las variables, excepto Cardiovascular diseases y 
db_cayafam <- readxl::read_excel("~/Desktop/Github_Proyectos/CAYAFAM-1/db_cayafam1_final.xlsx") 
```

We relabel some variables Sex and the period of exposure (Evento) for subset the data.

```{r Cleaning Database}
# Lo mismo de arriba pero para la base de datos completa 
#Sexo ("Hombre","Mujer")
db_cayafam <- db_cayafam %>%            
  dplyr::mutate(Sexo = factor(Sexo, levels = c(1,2),  #cambiamos los niveles    
                       labels = c("Hombre", "Mujer"))) %>%  
  dplyr::rename(Sexo = Sexo)

levels(db_cayafam$Sexo) #revisamos los nuevos niveles 

#Periodo de exposición  (evento) 
db_cayafam <- db_cayafam %>%            
  dplyr::mutate(Evento = factor(Evento, levels = c("medicion_invierno_arm_1","medicion_verano_arm_1"),  #cambiamos los niveles    
                                        labels = c("Invierno", "Verano"))) %>% 
  dplyr::rename(Evento = Evento)

levels(db_cayafam$Evento) #revisamos los nuevos niveles 
```

We made subset from our initial database (db_cayafam1)

```{r Subset, echo = FALSE}
#We do a subset for sex, period of exposure and for some environmental variables 
db_cayafam_summer <- subset(db_cayafam, Evento == "Verano") #contiene solo a hombres (N = 24) / ambos periodos (N = 48)
db_cayafam_winter <- subset(db_cayafam, Evento == "Invierno") #contiene solo a mujeres (N = 68) / ambos periodos (N = 136)
db_cayafam_env <- db_cayafam %>% dplyr::select(Evento, PM10_max:H_p75) #variables ambientales de interes / no se utiliza.
```

### Descriptive statistics

-   Table 1. Demographic and health characteristics of older adults.

```{r Table 1 Health Characteristics OA by Sex }
db_cayafam %>% distinct(Folio, Sexo) %>% 
               summarytools::freq(Sexo, report.nas = FALSE)

#Summary Age
db_cayafam_winter %>%  summarise(mean_Edad = round(mean(Edad), digits = 1), 
                                  sd_Edad = round(sd(Edad), digits = 1))

#Summary grouped by sex for Age
db_cayafam_winter %>% group_by(Sexo) %>% 
                      summarise(mean_Edad = round(mean(Edad), digits = 1), 
                                 sd_Edad = round(sd(Edad), digits = 1))

t.test(Edad ~ Sexo, data = db_cayafam_winter) #T.test Age - Sexo


#Summary Weight
db_cayafam_winter %>%  summarise(mean_Peso = round(mean(Peso, na.rm = TRUE),  digits = 1), 
                          sd_Peso = round(sd(Peso, na.rm = TRUE), digits = 1))

#Summary grouped by sex for Weight
db_cayafam_winter %>% group_by(Sexo) %>%  
                       summarise(mean_Weight = round(mean(Peso , na.rm = TRUE), digits = 1), 
                                 sd_Weight = round(sd(Peso, na.rm = TRUE), digits = 1))


t.test(Peso ~ Sexo, data = db_cayafam_winter) #T.test Weight - Sex

#Summary Height 
db_cayafam_winter %>%  summarise(mean_Height = round(mean(Talla, na.rm = TRUE), digits = 1), 
                                  sd_Height = round(sd(Talla, na.rm = TRUE), digits = 1))

#Summary grouped by sex for Height
db_cayafam_winter %>% group_by(Sexo) %>%  
                       summarise(mean_Height = round(mean(Talla , na.rm = TRUE), digits = 1), 
                                 sd_Height = round(sd(Talla, na.rm = TRUE), digits = 1))

t.test(Talla ~ Sexo, data = db_cayafam_winter) #T.test Height - Sex 


#Summary Obesity 
gmodels::CrossTable(db_cayafam_winter$Obesidad, db_cayafam_winter$Sexo, chisq = TRUE)


#Summary Waist 
db_cayafam %>%  summarise(mean_Waist = round(mean(CC , na.rm = TRUE), digits = 1), 
                                  sd_Waist = round(sd(CC, na.rm = TRUE), digits = 1))

#Summary grouped by sex for Waist
db_cayafam_winter %>% group_by(Sexo) %>%  
                       summarise(mean_Waist = round(mean(CC , na.rm = TRUE), digits = 1), 
                                 sd_Waist = round(sd(CC, na.rm = TRUE), digits = 1))

t.test(CC ~ Sexo, data = db_cayafam_winter) #T.test Waist - Sex

#Summary Smoking
gmodels::CrossTable(db_cayafam_winter$Smoking, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Diabetes
gmodels::CrossTable(db_cayafam_winter$Diabetes, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Hypertension
gmodels::CrossTable(db_cayafam_winter$HTA, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Respiratory diseases
gmodels::CrossTable(db_cayafam_winter$EnfResp, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Cardiovascular diseases
gmodels::CrossTable(db_cayafam_winter$Cardio_categ, db_cayafam_winter$Sexo, chisq = TRUE) # no encontrada

#Summary Multi-morbidity diseases
gmodels::CrossTable(db_cayafam_winter$Multimorbility, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Respiratory query last months 
gmodels::CrossTable(db_cayafam$consultas_respiratorias, db_cayafam$Sexo, chisq = TRUE) # no encontrada

#Summary Cardiovascular query last months 
gmodels::CrossTable(db_cayafam_winter$consultas_cardiacas, db_cayafam_winter$Sexo, chisq = TRUE) 

#Summary Hosptilization last months  (HospRespINV_ca2_14)
gmodels::CrossTable(db_cayafam_winter$HospRespINV_ca2_14, db_cayafam_winter$Sexo, chisq = TRUE)
```

We continue analizing de health characteristics and cardiorespiratory variables.

-   Table 2. Health characteristic and cardiorespiratory variables by High and Low Exposure.

```{r Table 2. Health Characteristic and Cardiorespiratory OA by Period of Exposure, message=FALSE, warning=FALSE, include=FALSE}

### Blood pressure ###

##Systolic blood pressure
db_cayafam %>%  
  dplyr::summarize(mean_PAS = round(mean(PAS_basal, na.rm = TRUE), digits = 1),  #Total period 
                              sd_PAS = round(sd(PAS_basal, na.rm = TRUE), digits = 1))


db_cayafam %>% filter(Evento == "medicion_invierno_arm_1") %>%  #High Exposure
  dplyr::summarize(mean_PAS = round(mean(PAS_basal, na.rm = TRUE), digits = 1), 
                              sd_PAS = round(sd(PAS_basal, na.rm = TRUE), digits = 1))


db_cayafam %>% filter(Evento == "medicion_verano_arm_1") %>%  #Low Exposure
  dplyr::summarize(mean_PAS = round(mean(PAS_basal, na.rm = TRUE), digits = 1), 
                              sd_PAS = round(sd(PAS_basal, na.rm = TRUE), digits = 1))

t.test(PAS_basal ~ Evento, data = db_cayafam) #T.test PAS_basal - Evento

#Diastolic blood pressure 
db_cayafam %>%  
  dplyr::summarize(mean_PAS = round(mean(PAD_basal, na.rm = TRUE), digits = 1),  #Total period 
                              sd_PAS = round(sd(PAD_basal, na.rm = TRUE), digits = 1))


db_cayafam %>% filter(Evento == "medicion_invierno_arm_1") %>%  #High Exposure
  dplyr::summarize(mean_PAS = round(mean(PAD_basal, na.rm = TRUE), digits = 1), 
                              sd_PAS = round(sd(PAD_basal, na.rm = TRUE), digits = 1))


db_cayafam %>% filter(Evento == "medicion_verano_arm_1") %>%  #Low Exposure
  dplyr::summarize(mean_PAD = round(mean(PAD_basal, na.rm = TRUE), digits = 1), 
                              sd_PAD = round(sd(PAD_basal, na.rm = TRUE), digits = 1))

t.test(PAD_basal ~ Evento, data = db_cayafam) #T.test PAD_basal - Evento

### Respiratory function ###
db_cayafam %>%  
  dplyr::summarize(mean_PEF = round(mean(PEF_max, na.rm = TRUE), digits = 1),  #Total period 
                              sd_PEF = round(sd(PEF_max, na.rm = TRUE), digits = 1))

db_cayafam_winter %>%  #High Exposure
  dplyr::summarize(mean_PEF = round(mean(PEF_max, na.rm = TRUE), digits = 1),   
                              sd_PEF = round(sd(PEF_max, na.rm = TRUE), digits = 1))

db_cayafam_summer %>% #Low Exposure
  dplyr::summarize(mean_PEF = round(mean(PEF_max, na.rm = TRUE), digits = 1),   
                              sd_PEF = round(sd(PEF_max, na.rm = TRUE), digits = 1))

t.test(PEF_max ~ Evento, data = db_cayafam) #T.test PEF_max - Evento

### Respiratory symptons ###
gmodels::CrossTable(db_cayafam$SintResp, db_cayafam$Evento, chisq = TRUE)

# Tos sin resfrio
gmodels::CrossTable(db_cayafam$tos_sin_resfrio, db_cayafam$Evento, chisq = TRUE )

# Flemas sin resfrio
gmodels::CrossTable(db_cayafam$flemas_sin_resfrio, db_cayafam$Evento, chisq = TRUE )

# Sibilancias
gmodels::CrossTable(db_cayafam$sibilancias, db_cayafam$Evento, chisq = TRUE )

### Physical Activity GPAQ ###
#Moderate intensity  (TOTAL_MIN_WEEK)
db_cayafam %>%  
  dplyr::summarize(mean = round(mean(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1),  #Total period 
                              sd = round(sd(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1))

db_cayafam_winter %>%  #High Exposure
  dplyr::summarize(mean = round(mean(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1),   
                              sd = round(sd(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1))

db_cayafam_summer %>% #Low Exposure
  dplyr::summarize(mean = round(mean(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1),   
                              sd = round(sd(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1))

t.test(TOTAL_MIN_WEEK ~ Evento, data = db_cayafam) #T.test TOTAL_MIN_WEEK - Evento

# >= 150 minutos of moderate intensity (Cumple_150minweek)
gmodels::CrossTable(db_cayafam$Cumple_150minweek, db_cayafam$Evento, chisq = TRUE)

# Level Physical Fitness (Pcondition)
gmodels::CrossTable(db_cayafam$Pcondition, db_cayafam$Evento, chisq = TRUE)

# si subimos el codigo, una opcion es realizar las tablas con esta libreria

# Table 1
library(kableExtra) 

# Table 2

library(kableExtra)

```

Then we make summary statistics for environmental variables for the both periods of exposure (High and Low)

-   Table 3. Air pollution and environmental variables by Period of Exposure

```{r Table 3. Air pollution and environmental variables by Period of Exposure, }

# no se hizo porque no tenemos las concentraciones de los contaminantes en el periodo (se pueden obtener desde el SINCA si es necesario)

```

### Multivariate Analysis

We create models to explain the variability of the different outcomes

#### Systolic Blood Pressure

For the systolic blood pressure models we use linear regression models to measure the association between the period of high exposure to pollutants and the outcome (SBP)

```{r Systolic Blood Pressure}

#Dejamos como categoria de referencia la variable de baja exposición (Verano).
db_cayafam$Evento <- relevel(db_cayafam$Evento, ref = "Verano")

#modelo ajustado

model1_PAS <- lm(formula = PAS_basal ~ factor(Evento)  + Edad + factor(Sexo) + factor(Smoking) + factor(Obesidad) + factor(Diabetes) + factor(HTA) + factor(Pcondition),  data = db_cayafam)

summary(model1_PAS)
confint(model1_PAS)

broom::tidy(model1_PAS)
broom::glance(model1_PAS)
broom::augment(model1_PAS)

# Generamos un grafico de coeficientes 
PAS_coef <- data.frame(
  exposure = c("High Exposure", "Physical Condition \n (Regular)", "Physical Condition \n (Optimal)" ),
  estimate = c(6.77, 2.34, 0.17),
  conf.low = c(1.04, -8.11, -11.53),
  conf.high = c(12.5, 12.79, 11.88))

#creamos el grafico usando ggplot
Fig.A_model_PAS <- ggplot2::ggplot(data = PAS_coef, mapping =aes( x= estimate, y = exposure)) + 
  geom_point() + 
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), shape = 16) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + 
  xlab("Beta estimates") + ylab("Exposures")  + theme_minimal(base_size = 12) + xlim(-15, 15)  + theme(plot.title = element_text (face='bold')) + ggtitle("A) Systolic BP") 

library(tidyverse)


```

#### Diastolic Blood Pressure

For the diastolic blood pressure models we use linear regression to measure the association between the period of high exposure to pollutants and the outcome (DBP)

```{r Diastolic Blood Pressure}

#Dejamos como categoria de referencia la variable de baja exposición (Verano).
db_cayafam$Evento <- relevel(db_cayafam$Evento, ref = "Verano")

#generamos el modelo 
model1_PAD <- lm(formula = PAD_basal ~ factor(Evento) + Edad + factor(Sexo) + factor(Smoking) + factor(Obesidad) + factor(Diabetes) + factor(HTA) + factor(Pcondition),  data = db_cayafam)

#resumen del modelo 
summary(model1_PAD)
confint(model1_PAD)

#ordenamos el modelo en tablas 
broom::tidy(model1_PAD)
broom::glance(model1_PAD)
broom::augment(model1_PAD)

# Generamos un grafico

PAD_coef <- data.frame(
  exposure = c("High Exposure", "Physical Condition \n (Regular)", "Physical Condition \n (Optimal)" ),
  estimate = c(3.51, 2.88, 2.64),
  conf.low = c(0.72, -2.18, -3.03),
  conf.high = c(6.29, 7.94, 8.32))


#creamos el grafico usando ggplot
Fig.B_model_PAD  <- ggplot2::ggplot(data = PAD_coef, mapping =aes( x= estimate, y = exposure)) + 
  geom_point() + 
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), shape = 16) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + 
  xlab("Beta estimates") + ylab("Exposures")  + theme_minimal(base_size = 12) + xlim(-15, 15)  + theme(plot.title = element_text (face='bold')) + ggtitle("B) Diastolic BP") 


```

#### Respiratory Symptoms

For the respiratory symptoms models we use logistic regression due to the outcome variable is binary, and it's categorize as a presence or absence of respiratory illness.

```{r Respiratory Symptoms}
# Algunas librerias 
library(MASS)

model1_EnfResp <- glm(formula = SintResp ~ factor(Evento) + Edad + factor(Sexo) + factor(Pcondition) + factor(Multimorbility) + factor(Smoking) + factor(Obesidad),  data = db_cayafam, family = binomial) 

summary(model1_EnfResp)

# exponenciamos los resultados para obtener nuestros OR
exp(cbind(OR = coef(model1_EnfResp), confint(model1_EnfResp)))

# Generamos un grafico de OR

EnfResp_OR <- data.frame(
  exposure = c("High Exposure", "Physical Condition \n (Regular)", "Physical Condition \n (Optimal)" ),
  estimate = c(4.43, 0.94, 1.17),
  conf.low = c(2.07, 0.25, 0.72),
  conf.high = c(10.04, 3.42, 5.15))

#creamos el grafico usando ggplot

Fig.C_model_SintResp <- ggplot2::ggplot(data = EnfResp_OR, mapping =aes( x= estimate, y = exposure)) + 
  geom_point() + 
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), shape = 16) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red")  + 
  scale_x_log10() + #cambiamos a escala logaritmica
  xlab("OR") + ylab("Exposures")  + theme_minimal(base_size = 12) + theme(plot.title = element_text (face='bold')) + ggtitle("C) Respiratory Symptoms") 



```

#### Lung Function

In the case of lung function we create

```{r Lung Function}

# construimos el modelo para la funcion pulmonar
model1_lungfuction <- lm(formula = PEF_max ~ factor(Evento) + Edad + factor(Sexo) + factor(Smoking) + factor(Obesidad) + factor(HTA) + factor(Pcondition),  data = db_cayafam)

#resumen del modelo 
summary(model1_lungfuction)
confint(model1_lungfuction)

#ordenamos el modelo en tablas 
broom::tidy(model1_lungfuction)
broom::glance(model1_lungfuction)
broom::augment(model1_lungfuction)


```


#### Figures 

We add a forest plot to show the changes in the beta coef estimates for the systolic and diastolic blood pressure, and the changes in the risk for the respiratory symptoms was measured trough odds ratio (OR). 

```{r Figure 1}

# we group all the figures for each of the models in one figure ussing the library patchwork. 

fig1 <- ((Fig.A_model_PAS / Fig.B_model_PAD) | Fig.C_model_SintResp / plot_spacer() ) + 
  plot_annotation(tag_levels = "A")

col1 <- (Fig.A_model_PAS / Fig.B_model_PAD) + plot_layout(heights = c(1, 1))
col2 <- (Fig.C_model_SintResp  / plot_spacer()) + plot_layout(heights = c(0.915, 1))

col1|col2

ggsave("fig_1_models.tiff", width = 10, height = 10, dpi = 400) 

```






#### Other visualizations / PRUEBAS (presentaciones, seminarios, congresos, webinars)

```{r Correlations}
install.packages("DataExplorer")
library(DataExplorer)

db_cayafam_cor <- db_cayafam %>% dplyr::select( PM10_max:NO2_max)%>% drop_na() 


#exploración normalidad de variales
DataExplorer::plot_histogram(db_cayafam_cor) #histogram de las variables
plot_histogram(db_cayafam_env) #altamente dispersas, quizas es necesario hacer una transformación para mejorar 

r_corr_matrix <- cor(db_cayafam_cor, method = "spearman")
r_corr_matrix

#correlacion con corrplot ----------------------
library(corrplot)
corrplot(r_corr_matrix, method ="shade", addCoef.col = FALSE)

#correlacion con ggplot -----------------------
cormat <- round(cor(db_cayafam_cor),2) #create a correlation
head(cormat)

library(reshape2)
melted_cormat <-melt(cormat) #we melt the correlation matrix
head(melted_cormat)

#creamos con ggplot corplot
install.packages("viridis")
library(viridis)
corplot <- ggplot(data=melted_cormat, aes(x=Var1, y =Var2, fill= value)) + geom_tile() + scale_fill_viridis_c(name = "Spearman \n Correlation") + theme_minimal() 


#funciones para obtener solo la diagnal 
# Get lower triangle of the correlation matrix
  get_lower_tri<-function(cormat){
    cormat[upper.tri(cormat)] <- NA
    return(cormat)
  }
  # Get upper triangle of the correlation matrix
  get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
  }
  
# nuevo grafico con ggplot
  
upper_tri <- get_upper_tri(cormat)

library(reshape2)
melted_cormat <- melt(upper_tri, na.rm = TRUE)

ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_viridis_c(name = "Spearman \n Correlation") +
  theme_minimal() + geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  legend.justification = c(1, 0),
  legend.position = c(0.6, 0.7),
  legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                title.position = "top", title.hjust = 0.5))

```

```{r adicionales}

### ejemplo grafico Env Factors --------
pm10 <- ggboxplot(db_cayafam, x = "Evento", y = "PM10_max", 
          color = "Evento", palette = c("#00AFBB", "#E7B800"),
          order = c("Invierno", "Verano"),
          ylab = "PM10", xlab = "") 

pm25 <- ggboxplot(db_cayafam1, x = "Evento", y = "PM2.5_max", 
          color = "Evento", palette = c("#00AFBB", "#E7B800"),
          order = c("Invierno", "Verano"),
          ylab = "PM2.5", xlab = "Periodo de Exposición") 

no2 <- ggboxplot(db_cayafam1, x = "Evento", y = "NO2_max", 
          color = "Evento", palette = c("#00AFBB", "#E7B800"),
          order = c("Invierno", "Verano"),
          ylab = "NO2", xlab = "") 


pm10 + pm25 + no2



### Comparación de modelos ----------
install.packages("performance")
library(performance)

(comp <- compare_performance(m1_RS, m2_RS, m3_RS, modelo_prueba))

install.packages("see")
plot(comp)

```

```{r}

```

\`\`\`
