---
title: ""
author: "Sandra Cortes, Alan Domínguez"
date: "7/23/2021"
output: html_document
---

## CAYAFAM 1

The objective of this study was to analyze the association between air pollution and cardiorespiratory variables, considering the effect of physical condition, in OA that live in areas with high levels of air pollution in the Metropolitan Region of Chile.

```{r Libraries, include=FALSE}
library(tidyverse)
library(skimr)
library(haven)
library(readxl)
library(patchwork)
library(gmodels)
library(summarytools)
library(ggpubr)
library(ggthemes)
library(DataExplorer)

```

To charge the database from SPSS we use the function read_sav from the library(haven) for the initial database, and the function read_excel from the library(readxl).

```{r Loading Database, echo = FALSE}
db_cayafam1 <- readxl::read_excel("~/Desktop/Github_Proyectos/CAYAFAM-1/db_cayafam1.xlsx") #base de datos con 54 variables (Descriptiva tabla 1)
db_cayafam1_vf <- readxl::read_excel("~/Desktop/Github_Proyectos/CAYAFAM-1/db_cayafam1_vf.xlsx") 
db_cayafam <- readxl::read_excel("~/Desktop/Github_Proyectos/CAYAFAM-1/db_cayafam1_final.xlsx") #con esta la mayoria 
```

We relabel some variables Sex and the period of exposure (Evento) for subset the data.

```{r Cleaning Database}
# Lo mismo de arriba pero para la base de datos completa 
#Sexo ("Hombre","Mujer")
db_cayafam <- db_cayafam %>%            
  dplyr::mutate(Sexo = factor(Sexo, levels = c(1,2),  #cambiamos los niveles    
                       labels = c("Hombre", "Mujer"))) %>%  
  dplyr::rename(Sexo = Sexo)

levels(db_cayafam$Sexo) #revisamos los nuevos niveles 

#Periodo de exposición  (evento) 
db_cayafam <- db_cayafam %>%            
  dplyr::mutate(Evento = factor(Evento, levels = c("medicion_invierno_arm_1","medicion_verano_arm_1"),  #cambiamos los niveles    
                                        labels = c("Invierno", "Verano"))) %>% 
  dplyr::rename(Evento = Evento)

levels(db_cayafam$Evento) #revisamos los nuevos niveles 
```

We made subset from our initial database (db_cayafam1)

```{r Subset, echo = FALSE}
#We do a subset for sex, period of exposure and for some environmental variables 
db_cayafam_summer <- subset(db_cayafam, Evento == "Verano") #contiene solo a hombres (N = 24) / ambos periodos (N = 48)
db_cayafam_winter <- subset(db_cayafam, Evento == "Invierno") #contiene solo a mujeres (N = 68) / ambos periodos (N = 136)
db_cayafam_env <- db_cayafam %>% dplyr::select(Evento, PM10_max:H_p75) #variables ambientales de interes
```

### Descriptive statistics

-   Table 1. Demographic and health characteristics of older adults.

```{r Table 1 Health Characteristics OA by Sex }
db_cayafam %>% distinct(Folio, Sexo) %>% 
               summarytools::freq(Sexo, report.nas = FALSE)

#Summary Age
db_cayafam_winter %>%  summarise(mean_Edad = round(mean(Edad), digits = 1), 
                                  sd_Edad = round(sd(Edad), digits = 1))

#Summary grouped by sex for Age
db_cayafam_winter %>% group_by(Sexo) %>% 
                      summarise(mean_Edad = round(mean(Edad), digits = 1), 
                                 sd_Edad = round(sd(Edad), digits = 1))

t.test(Edad ~ Sexo, data = db_cayafam_winter) #T.test Age - Sexo


#Summary Weight
db_cayafam_winter %>%  summarise(mean_Peso = round(mean(Peso, na.rm = TRUE),  digits = 1), 
                          sd_Peso = round(sd(Peso, na.rm = TRUE), digits = 1))

#Summary grouped by sex for Weight
db_cayafam_winter %>% group_by(Sexo) %>%  
                       summarise(mean_Weight = round(mean(Peso , na.rm = TRUE), digits = 1), 
                                 sd_Weight = round(sd(Peso, na.rm = TRUE), digits = 1))


t.test(Peso ~ Sexo, data = db_cayafam1_winter) #T.test Weight - Sex

#Summary Height 
db_cayafam1_winter %>%  summarise(mean_Height = round(mean(Talla, na.rm = TRUE), digits = 1), 
                                  sd_Height = round(sd(Talla, na.rm = TRUE), digits = 1))

#Summary grouped by sex for Height
db_cayafam1_winter %>% group_by(Sexo) %>%  
                       summarise(mean_Height = round(mean(Talla , na.rm = TRUE), digits = 1), 
                                 sd_Height = round(sd(Talla, na.rm = TRUE), digits = 1))

t.test(Talla ~ Sexo, data = db_cayafam1_winter) #T.test Height - Sex 


#Summary Obesity 
gmodels::CrossTable(db_cayafam1_winter$`Obsidad(1/0)`, db_cayafam1_winter$Sexo, chisq = TRUE)


#Summary Waist 
db_cayafam1 %>%  summarise(mean_Waist = round(mean(Perimetro_cintura , na.rm = TRUE), digits = 1), 
                                  sd_Waist = round(sd(Perimetro_cintura, na.rm = TRUE), digits = 1))

#Summary grouped by sex for Waist
db_cayafam1_winter %>% group_by(Sexo) %>%  
                       summarise(mean_Waist = round(mean(Perimetro_cintura , na.rm = TRUE), digits = 1), 
                                 sd_Waist = round(sd(Perimetro_cintura, na.rm = TRUE), digits = 1))

t.test(Perimetro_cintura ~ Sexo, data = db_cayafam1_winter) #T.test Waist - Sex

#Summary Smoking
gmodels::CrossTable(db_cayafam1_winter$`Fuma(si/no)`, db_cayafam1_winter$Sexo, chisq = TRUE) #eliminar luego
gmodels::CrossTable(db_cayafam_winter$Smoking, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Diabetes
gmodels::CrossTable(db_cayafam_winter$Diabetes, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Hypertension
gmodels::CrossTable(db_cayafam_winter$HTA, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Respiratory diseases
gmodels::CrossTable(db_cayafam_winter$EnfResp, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Cardiovascular diseases
gmodels::CrossTable(db_cayafam$EnfResp, db_cayafam$Sexo, chisq = TRUE)#Periodo de alta exposición (invierno)

gmodels::CrossTable(db_cayafam1_summer$Enf_cardio, db_cayafam1_summer$Sexo, chisq = TRUE) #Periodo de baja exposcición (verano)

#Summary Multi-morbidity diseases
gmodels::CrossTable(db_cayafam_winter$Multimorbility, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Respiratory query last months 
gmodels::CrossTable(db_cayafam$consultas_respiratorias, db_cayafam$Sexo, chisq = TRUE)

#Summary Cardiovascular query last months 
gmodels::CrossTable(db_cayafam_winter$consultas_cardiacas, db_cayafam_winter$Sexo, chisq = TRUE)

#Summary Hosptilization last months  (HospRespINV_ca2_14)
gmodels::CrossTable(db_cayafam_winter$HospRespINV_ca2_14, db_cayafam_winter$Sexo, chisq = TRUE)
```

We continue analizing de health characteristics and cardiorespiratory variables.

-   Table 2. Health characteristic and cardiorespiratory variables by High and Low Exposure.

```{r Table 2. Health Characteristic and Cardiorespiratory OA by Period of Exposure, message=FALSE, warning=FALSE, include=FALSE}

### Blood pressure ###

##Systolic blood pressure
db_cayafam %>%  
  dplyr::summarize(mean_PAS = round(mean(PAS_basal, na.rm = TRUE), digits = 1),  #Total period 
                              sd_PAS = round(sd(PAS_basal, na.rm = TRUE), digits = 1))


db_cayafam %>% filter(Evento == "medicion_invierno_arm_1") %>%  #High Exposure
  dplyr::summarize(mean_PAS = round(mean(PAS_basal, na.rm = TRUE), digits = 1), 
                              sd_PAS = round(sd(PAS_basal, na.rm = TRUE), digits = 1))


db_cayafam %>% filter(Evento == "medicion_verano_arm_1") %>%  #Low Exposure
  dplyr::summarize(mean_PAS = round(mean(PAS_basal, na.rm = TRUE), digits = 1), 
                              sd_PAS = round(sd(PAS_basal, na.rm = TRUE), digits = 1))

t.test(PAS_basal ~ Evento, data = db_cayafam) #T.test PAS_basal - Evento

#Diastolic blood pressure 
db_cayafam %>%  
  dplyr::summarize(mean_PAS = round(mean(PAD_basal, na.rm = TRUE), digits = 1),  #Total period 
                              sd_PAS = round(sd(PAD_basal, na.rm = TRUE), digits = 1))


db_cayafam %>% filter(Evento == "medicion_invierno_arm_1") %>%  #High Exposure
  dplyr::summarize(mean_PAS = round(mean(PAD_basal, na.rm = TRUE), digits = 1), 
                              sd_PAS = round(sd(PAD_basal, na.rm = TRUE), digits = 1))


db_cayafam %>% filter(Evento == "medicion_verano_arm_1") %>%  #Low Exposure
  dplyr::summarize(mean_PAD = round(mean(PAD_basal, na.rm = TRUE), digits = 1), 
                              sd_PAD = round(sd(PAD_basal, na.rm = TRUE), digits = 1))

t.test(PAD_basal ~ Evento, data = db_cayafam) #T.test PAD_basal - Evento

### Respiratory function ###
db_cayafam %>%  
  dplyr::summarize(mean_PEF = round(mean(PEF_max, na.rm = TRUE), digits = 1),  #Total period 
                              sd_PEF = round(sd(PEF_max, na.rm = TRUE), digits = 1))

db_cayafam1 %>% filter(Evento == "Invierno") %>%  #High Exposure
  dplyr::summarize(mean_PEF = round(mean(PEF_max, na.rm = TRUE), digits = 1),   
                              sd_PEF = round(sd(PEF_max, na.rm = TRUE), digits = 1))

db_cayafam1 %>% filter(Evento == "Verano") %>%  #Low Exposure
  dplyr::summarize(mean_PEF = round(mean(PEF_max, na.rm = TRUE), digits = 1),   
                              sd_PEF = round(sd(PEF_max, na.rm = TRUE), digits = 1))

t.test(PEF_max ~ Evento, data = db_cayafam1) #T.test PEF_max - Evento

### Respiratory symptons ###
gmodels::CrossTable(db_cayafam$SintResp, db_cayafam$Evento, chisq = TRUE)

# Tos sin resfrio
gmodels::CrossTable(db_cayafam$tos_sin_resfrio, db_cayafam$Evento, chisq = TRUE )

# Flemas sin resfrio
gmodels::CrossTable(db_cayafam$flemas_sin_resfrio, db_cayafam$Evento, chisq = TRUE )

# Sibilancias
gmodels::CrossTable(db_cayafam$sibilancias, db_cayafam$Evento, chisq = TRUE )

### Physical Activity GPAQ ###
#Moderate intensity  (TOTAL_MIN_WEEK)
db_cayafam %>%  
  dplyr::summarize(mean = round(mean(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1),  #Total period 
                              sd = round(sd(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1))

db_cayafam %>% filter(Evento == "medicion_invierno_arm_1") %>%  #High Exposure
  dplyr::summarize(mean = round(mean(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1),   
                              sd = round(sd(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1))

db_cayafam %>% filter(Evento == "medicion_verano_arm_1") %>%  #Low Exposure
  dplyr::summarize(mean = round(mean(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1),   
                              sd = round(sd(TOTAL_MIN_WEEK, na.rm = TRUE), digits = 1))

t.test(TOTAL_MIN_WEEK ~ Evento, data = db_cayafam) #T.test TOTAL_MIN_WEEK - Evento

# >= 150 minutos of moderate intensity (Cumple_150minweek)
gmodels::CrossTable(db_cayafam$Cumple_150minweek, db_cayafam$Evento, chisq = TRUE )

# Level Physical Fitness (Pcondition)
gmodels::CrossTable(db_cayafam$Pcondition, db_cayafam$Evento, chisq = TRUE )

```

Then we make summary statistics for environmental variables for the both periods of exposure (High and Low)

-   Table 3. Air pollution and environmental variables by Period of Exposure

```{r Table 3. Air pollution and environmental variables by Period of Exposure, }




```

### Multivariate Analysis

We create models to explain the variability of the different outcomes

-   Respiratory Symptoms model

    For the respiratory symptoms models we use logistic regression due to the outcome variable is binary, and it's categorize as a presence or absence of respiratory illness.

```{r Respiratory Symptoms}
# Algunas librerias 
library(MASS)

model1_EnfResp <- glm(formula = SintResp ~ factor(Evento) + Edad + factor(Sexo) + factor(Pcondition) + factor(Multimorbility) + factor(Smoking) + factor(Obesidad),  data = db_cayafam, family = binomial) 

summary(model1_EnfResp)

# exponenciamos los resultados para obtener nuestros OR
exp(cbind(OR = coef(model1_EnfResp), confint(model1_EnfResp)))

# Generamos un grafico de OR

EnfResp_OR <- data.frame(
  exposure = c("High Exposure", "Physical Condition (Regular)", "Physical Condition  (Optimal)" ),
  estimate = c(4.43, .94, 1.17),
  conf.low = c(2.07, .25, .72),
  conf.high = c(10.04, 3.42, 5.15))

#creamos el grafico usando ggplot

Fig.C_model_SintResp <- ggplot2::ggplot(data = EnfResp_OR, mapping =aes( x= estimate, y = exposure)) + 
  geom_point() + 
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), shape = 16) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red")  + 
  scale_x_log10() + #cambiamos a escala logaritmica
  xlab("OR adjusted CI95%") + ylab("Exposures")  + theme_minimal() + ggtitle("C) Respiratory Symptoms") + theme(plot.title = element_text (face='bold')) 





```

```{r Systolic Blood Pressure}

#Dejamos como categoria de referencia la variable de baja exposición (Verano).
db_cayafam$Evento <- relevel(db_cayafam$Evento, ref = "Verano")

#modelo ajustado
model1_PAS <- lm(formula = PAS_basal ~ factor(Evento)  + Edad + factor(Sexo) + factor(Smoking) + factor(Obesidad) + factor(Diabetes) + factor(HTA) + factor(Pcondition),  data = db_cayafam)

summary(model1_PAS)
confint(model1_PAS)


broom::tidy(model1_PAS)
broom::glance(model1_PAS)
broom::augment(model1_PAS)

# Generamos un grafico de coeficientes beta

PAS_coef <- data.frame(
  exposure = c("High Exposure", "Physical Condition (Regular)", "Physical Condition  (Optimal)" ),
  estimate = c(6.77, 2.34, 0.17),
  conf.low = c(1.04, -8.11, -11.53),
  conf.high = c(12.5, 12.79, 11.88))

#creamos el grafico usando ggplot
Fig.A_model_PAS <- ggplot2::ggplot(data = PAS_coef, mapping =aes( x= estimate, y = exposure)) + 
  geom_point() + 
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), shape = 16) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + 
  xlab("Beta estimates") + ylab("Exposures")  + theme_minimal() + xlim(-15, 15) + ggtitle("A) Diastolic Blood Pressure") + theme(plot.title = element_text (face='bold')) 



```

```{r Diastolic Blood Pressure}

#Dejamos como categoria de referencia la variable de baja exposición (Verano).
db_cayafam$Evento <- relevel(db_cayafam$Evento, ref = "Verano")

#generamos el modelo 
model1_PAD <- lm(formula = PAD_basal ~ factor(Evento) + Edad + factor(Sexo) + factor(Smoking) + factor(Obesidad) + factor(Diabetes) + factor(HTA) + factor(Pcondition),  data = db_cayafam)

#resumen del modelo 
summary(model1_PAD)
confint(model1_PAD)

#ordenamos el modelo en tablas 
broom::tidy(model1_PAD)
broom::glance(model1_PAD)
broom::augment(model1_PAD)

# Generamos un grafico

PAD_coef <- data.frame(
  exposure = c("High Exposure", "Physical Condition (Regular)", "Physical Condition  (Optimal)" ),
  estimate = c(3.51, 2.88, 2.64),
  conf.low = c(.72, -2.18, -3.03),
  conf.high = c(6.29, 7.94, 8.32))

#creamos el grafico usando ggplot
Fig.B_model_PAS  <- ggplot2::ggplot(data = PAD_coef, mapping =aes( x= estimate, y = exposure)) + 
  geom_point() + 
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), shape = 16) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + 
  xlab("Beta estimates") + ylab("Exposures")  + theme_minimal() + xlim(-15, 15) + ggtitle("B) Systolic Blood Pressure") + theme(plot.title = element_text (face='bold')) 


Fig.A_model_PAS + 
Fig.B_model_PAS + 
Fig.C_model_SintResp




```

-   ideas de visualización

```{r Correlations}
install.packages("DataExplorer")
library(DataExplorer)

db_cayafam_cor <- db_cayafam %>% dplyr::select( PM10_max:NO2_max)%>% drop_na() 


#exploración normalidad de variales
DataExplorer::plot_histogram(db_cayafam_cor) #histogram de las variables
plot_histogram(db_cayafam_env) #altamente dispersas, quizas es necesario hacer una transformación para mejorar 

r_corr_matrix <- cor(db_cayafam_cor, method = "spearman")
r_corr_matrix

#correlacion con corrplot ----------------------
library(corrplot)
corrplot(r_corr_matrix, method ="shade", addCoef.col = FALSE)

#correlacion con ggplot -----------------------
cormat <- round(cor(db_cayafam_cor),2) #create a correlation
head(cormat)

library(reshape2)
melted_cormat <-melt(cormat) #we melt the correlation matrix
head(melted_cormat)

#creamos con ggplot corplot
install.packages("viridis")
library(viridis)
corplot <- 0





```

```{r adicionales}

### ejemplo grafico Env Factors --------
pm10 <- ggboxplot(db_cayafam1, x = "Evento", y = "PM10_max", 
          color = "Evento", palette = c("#00AFBB", "#E7B800"),
          order = c("Invierno", "Verano"),
          ylab = "PM10", xlab = "") 

pm25 <- ggboxplot(db_cayafam1, x = "Evento", y = "PM2.5_max", 
          color = "Evento", palette = c("#00AFBB", "#E7B800"),
          order = c("Invierno", "Verano"),
          ylab = "PM2.5", xlab = "Periodo de Exposición") 

no2 <- ggboxplot(db_cayafam1, x = "Evento", y = "NO2_max", 
          color = "Evento", palette = c("#00AFBB", "#E7B800"),
          order = c("Invierno", "Verano"),
          ylab = "NO2", xlab = "") 


pm10 + pm25 + no2



### Comparación de modelos ----------
install.packages("performance")
library(performance)

(comp <- compare_performance(m1_RS, m2_RS, m3_RS, modelo_prueba))

install.packages("see")
plot(comp)

```

```{r}

```

\`\`\`
